Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-09-06 paper/ms.tex
Minor edits of appendix

Rather than using $\beta$ and its complex conjugate as components of the kernel, we instead
find the real and imaginary components of the complex equation for $\beta$ by taking the
sum and difference of the $\beta$ and $\beta^*$ equations (divided by 2 and $2i$, respectively), 
and then discard the $\beta^*$ equation, which is redundant.  This keeps the same number of
additional equations for each $\beta_j$ with non-zero imaginary component (four additional
equations in the extended matrix), but converts the complex equations to real, thus avoiding
complex arithmetic, while the kernel components with real $\beta_j$ still only have two 
additional components in the extended matrix.  The extended matrix adds an extra band above
and below the diagonal when the $\beta$ and $\beta^*$ equations are combined since this
mixes the components of the original equations which are shifted by one column from one
another.  We will enumerate the number of real $\beta$'s as $p_0$, while the total number 
of $\beta$'s is still $p$.  This gives a total number of equations to be solved with the 
extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $n$ is the number of data 
points.

For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric 
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $p$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
to be the real values of $\beta$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation}
% See 9/2/2016 notes
\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
\end{equation}
for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61).

Equation (60) has real and imaginary components:
\begin{eqnarray}
\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
\end{eqnarray}
with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran2015}.

Finally, equation (59) has real and imaginary components:
\begin{align}
-&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
 &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
\end{align}
with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) as well for these, which reduces the total number
of equations to solve.

With these additional equations, we can solve for the likelihood of sums of full Lorentzian kernels 
in $O(N)$ operations, but with slightly more computational expense due to having 4 extra variables 
for each Lorentzian with non-zero $\omega$, and due to having a slightly larger bandwidth due to the 
cross terms between the real and imaginary variables.  The off-diagonal components are $2(p+1)$ above 
and below the diagonal, for a total bandwidth of $4p+5$; in the case in which $p_0=p$ so that the
imaginary components of $\beta$ are zero, the off-diagonal components revert to $p+1$, giving
a bandwidth of $2p+3$.

As with the GRP method, we can buid an extended matrix which is real and symmetric, and which may be 
solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix 
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a 
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with real values of 
$\beta$ first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
x_2, ..., x_N]$.

>>>>

Rather than using $\beta$ and its complex conjugate as components of the kernel, we instead
find the real and imaginary components of the complex equation for $\beta$ by taking the
sum and difference of the $\beta$ and $\beta^*$ equations (divided by 2 and $2i$, respectively), 
and then discard the $\beta^*$ equation, which is redundant.  This keeps the same number of
additional equations for each $\beta_j$ with non-zero imaginary component (four additional
equations in the extended matrix), but converts the complex equations to real, thus avoiding
complex arithmetic, while the kernel components with real $\beta_j$ still only have two 
additional components in the extended matrix.  The extended matrix adds an extra band above
and below the diagonal when the $\beta$ and $\beta^*$ equations are combined since this
mixes the components of the original equations which are shifted by one column from one
another.  We will enumerate the number of real $\beta$'s as $p_0$, while the total number 
of $\beta$'s is still $p$.  This gives a total number of equations to be solved with the 
extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $N$ is the number of data 
points.

For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric 
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $p$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
to be the real values of $\beta$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
\end{equation}
for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61)
in \citet{Ambikasaran2015}.

Equation (60) in \citet{Ambikasaran2015} has real and imaginary components:
\begin{eqnarray} \label{Amb60}
\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
\end{eqnarray}
with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran2015}.

Finally, equation (59) in \citet{Ambikasaran2015} has real and imaginary components:
\begin{align} \label{Amb59}
-&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
 &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
\end{align}
with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) in \citet{Ambikasaran2015} as well for these, which reduces the total number
of equations to solve.

With these additional equations, we can solve for the likelihood of sums of full Lorentzian kernels 
in $O(N)$ operations, but with slightly more computational expense due to having 4 extra variables 
for each Lorentzian with non-zero $\omega$, and due to having a slightly larger bandwidth due to the 
cross terms between the real and imaginary variables.  The off-diagonal components are $p_0+2(p-p_0)+2$ above 
and below the diagonal, for a total bandwidth of $2p_0+4(p-p_0)+5$. In the case in which $p_0=p$, so 
that the imaginary components of each $\beta$ are zero, the off-diagonal components revert to $p+1$ 
above and below the diagonal, giving a bandwidth of $2p+3$ (the same as the GRP algorithm).

As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be 
solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix 
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a 
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with real values of 
$\beta$ first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [b_1/2, \{0; j=1,...,p_0+2(p-p_0)\}, b_2/2, ..., b_N/2]$.
The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2p_0+4(p-p_0)+5) \times N_{ex}$.
The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
with a banded algorithm.
